---
title: 'Reproducible Research: Peer Assignment 1'
author: "cws"
date: "2025-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment

This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.

### Loading and Preprocessing the Data

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Load the data
data <- read.csv("activity.csv")

# Process/transform the data (convert date to Date type)
data$date <- as.Date(data$date, format="%Y-%m-%d")

# View the data
head(data)
```

### Transform the Data into a suitable format
```{r}
activity_NONA <- na.omit(data)
head(activity_NONA)
```

### What are the mean and median steps taken per day
```{r}
# Aggregate steps per day for visualization
activity_per_day <- activity_NONA %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))

head(activity_per_day)
```

### Plot a histogram of the total steps per day
```{r}
ggplot(activity_per_day, aes(x = total_steps)) +
  geom_histogram(binwidth = 500, fill = 'steelblue3', color = 'black') +
  ggtitle("Total Number of Steps Taken Each Day") +
  xlab("Total Steps") +
  ylab("Frequency")
```

### Summarize the statistics of total steps per day
```{r}
summary(activity_per_day)
```

### Aggregate average steps per interval for visualization
```{r}
activity_per_interval <- activity_NONA %>%
  group_by(interval) %>%
  summarise(mean_steps = mean(steps))
```

### Plot the time series of average steps per interval
```{r}
ggplot(activity_per_interval, aes(x = interval, y = mean_steps)) + 
  geom_line(color = "blue") +
  ggtitle("Average Number of Steps Taken per Interval") +
  xlab("Interval") +
  ylab("Average Steps")
```

### Find the interval with the maximum average steps
```{r}
max_interval <- activity_per_interval[which.max(activity_per_interval$mean_steps),]
print(max_interval)
```

### Count the number of missing values in the data set
```{r}
missing_values <- colSums(is.na(data))
print(missing_values)
```

### Impute missing values using the mean steps per interval
```{r}
activity_filled <- data %>%
  left_join(activity_per_interval, by = "interval") %>%
  mutate(steps = ifelse(is.na(steps), mean_steps, steps)) %>%
  select(-mean_steps)  # Remove the extra column created by join

head(activity_filled)
```

### Aggregate total steps per day after filling missing values
```{r}
activity_filled2 <- activity_filled %>%
  group_by(date) %>%
  summarise(total_steps = sum(steps))
```

### Plot histogram after imputing missing values
```{r}
ggplot(activity_filled2, aes(x = total_steps)) +
  geom_histogram(binwidth = 500, fill = 'red', color = 'black') +
  ggtitle("Total Number of Steps Taken Each Day - Missing Values Added") +
  xlab("Total Steps") +
  ylab("Frequency")
```

### Summary statistics after imputing missing values
```{r}
summary(activity_filled2)
```

### Categorize days as weekday or weekend
```{r}
activity_filled <- activity_filled %>%
  mutate(weekdayType = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
```

### Compute mean steps per interval for weekdays vs weekends
```{r}
activity_filled3 <- activity_filled %>%
  group_by(interval, weekdayType) %>%
  summarise(mean_steps = mean(steps), .groups = "drop")

head(activity_filled3)
```

### Plot time series average steps by interval for weekdays and weekends
```{r}
ggplot(activity_filled3, aes(x = interval, y = mean_steps)) + 
  geom_line(color = "darkgreen") +
  facet_grid(rows = vars(weekdayType)) +
  ggtitle("Average Steps per Interval - Weekday vs. Weekend") +
  xlab("Interval") +
  ylab("Average Steps")
```

